A statistical framework for quantifying the nuclear export rate of influenza viral mRNAs

Version 1

Abstract

Abstract

Influenza A virus acutely transcribes viral mRNAs from the eight segmented viral genome when it infects. The kinetics of viral transcription, nuclear export of viral transcripts, and their potential variation between the eight segments are poorly characterised. Here we introduce a statistical framework for estimating the nuclear export rate of each segment from a snapshot of mRNA in situ localisation at a single time point, exploiting the cell-to-cell variation observed by an imaging-based in situ transcriptome assay. A mathematical modelling indicated that the relationship between the nuclear ratio and the total count of mRNAs in single cells is dictated by a proxy for the nuclear export rate. Using this model, we showed that the two influenza viral antigens hemagglutinin and neuraminidase were the slowest segments in the nuclear export, suggesting that influenza A virus uses the nuclear retention of viral transcripts to delay the expression of antigenic molecules. Our framework presented in this study can be widely used for investigating the nuclear retention of nascent transcripts produced in a transcription burst.

Introduction

Temporal gene expression is of vital importance for viral pathogens to proliferate in the host. Human exogenous retroviruses insert their RNA genome into the host chromatin and use host factors to minimise the viral protein production to evade the immune response and persist in the host 1. In contrast, influenza A virus acutely infects the respiratory tract and produces thousands of viral progenitors in a single round of infection to spread in the surrounding epithelial cells by transcribing and replicating its RNA genome 2. Influenza A virus genome is segmented into eight single-stranded RNAs, and expression of each gene is essential for productive infection 3. Unlike the human exogenous retroviruses, influenza A virus does not seem to have a sophisticated mechanism of controlling the viral gene expression at the transcription level. However, it has been known for decades that hemagglutinin (HA) and neuraminidase (NA), the two influenza viral antigens targeted by the host immune response, are produced late during the course of infection 4,5,6, a mechanism of which is currently unclear.

Eight viral RNA segments are transcribed by the same set of molecular machinery 7. Viral genome is contained in the viral envelope in association with viral proteins. The 5′ and 3′ ends of each viral genomic RNA hybridise to form an intramolecular loop 8. The RNA-dependent RNA polymerase of influenza A virus, comprising PB1 (polymerase basic 1), PB2 (polymerase basic 2), and PA (polymerase acidic) subunits, is associated with the 5′ and 3′ joint end; the rest of the loop is bound by the oligomeric viral nucleoprotein (NP) 9. The viral ribonucleoprotein complex (vRNPs) initiates the viral transcription in the nucleus by deriving a 5′-capped short RNA fragment from the host nascent RNA on the stalled RNA polymerase II (Pol II) 10. At the end of transcription, the viral polymerase adds a poly-A tail to the viral mRNA by stuttering on a short uridine repeat in the viral RNA template 11. Thus, eight viral segments sharing the common mechanism of viral transcription with no additional layers of transcriptional controls e.g., transcription-factor binding, the transcription of eight influenza viral genes is likely to occur at the same rate across the segments that is determined by the collision of the vRNP and the stalled Pol II.

Another possible factor that would determine the rate of transcription is the length of viral RNA template 12, 13, given that the duration of transcription is rate-limiting. However, this does not agree with the order of protein expression in the influenza virus infection: PB2 (2.4 kb), PB1 (2.3 kb) and PA (2.2 kb), the three longest viral genes, and NP (1.8 kb) are expressed first, followed by the expression of three segments HA (1.8 kb), NA (1.5 kb) and M (1.0 kb) 4, 6.

A third possibility is the nuclear retention of viral mRNAs. Nuclear retention of mRNAs has gained an attention as a point of control on the protein expression in the context of transcription burst 14. Previous studies reported the accumulation of M mRNAs in the nuclear speckles 15, NXF1-dependent nuclear export of HA, NA and M mRNAs 16, 17, and the nuclear export inhibition by the host factor hnRNPAB for M, HA, NP and NS mRNAs 18. It is becoming clear that the influenza viral segments show differential dependencies on the nuclear export factors; and this dependency is not accounted for by the RNA structure (i.e., whether they are intron-less, unspliced or spliced). Because an incomplete set of viral segments was studied in each of these studies, the influenza viral mRNA transport remains elusive.

In this paper, we introduce a statistical framework for systematically quantifying the nuclear transport across all the eight viral segments. We localised eight viral mRNAs simultaneously in single cells in situ. Intracellular distribution of viral mRNAs indicated that the nuclear export rate of viral transcripts varied according to the segments. A statistical model, exploiting cell-to-cell variation in the abundance of transcripts due to the stochastic nature of multiple virological processes prior to the onset of viral transcription, allowed for the estimation of nuclear export rate from the intracellular distribution of mRNAs observed at a single time point. With this model, we show that hemagglutinin and neuraminidase mRNAs were the slowest in exiting the nucleus, suggesting that influenza A virus uses nuclear retention of mRNAs to delay the production of antigenic molecules in the course of viral infection. This study demonstrates that the nuclear export rate of nascent transcripts can be estimated from a snapshot of RNA in situ measurement in single cells, and our result serves as a reference for the nuclear export rate of viral transcripts across the segments, shedding light on the comprehensive understanding on the nuclear export of viral transcripts.

Results

Diffraction-limited, single-colour imaging of viral mRNAs

First, we used single-molecule RNA FISH (smFISH) to study the viral transcription of PB1, HA, NP and NS. The lung carcinoma cell line A549 was infected with the WSN strain of influenza A virus at the multiplicity of infection (M.O.I.) of ∼2 for two hours, and the viral mRNAs were visualised with a tile of short oligonucleotide probes, each labelled with a single fluorophore 19 (Fig. 1A). Viral mRNAs were detected as single diffraction-limited spots in the three-dimensional image stacks, allowing for absolute mRNA quantification (Fig. 1B).

Viral mRNA detection by single-colour RNA FISH.

(A) A schematic diagram of mRNA labelling for the diffraction-limited imaging. Each mRNA was hybridised with a tile of oligonucleotide, each labelled with a single fluorophore at the 3′ end. (B) A representative field-of-view of the microscope image in which NP mRNAs were stained with Quasar 670. White, smFISH signal; blue, nucleus stained with DAPI. The red square indicates the area for the sequential close-up of the image. The signal intensity is shown on the same grey scale across the three panels, indicated on the right. (C) Spot detection in 3D. Spots are projected to either the XZ or YZ plane to indicate the distribution of spots in a typical 3D cell volume. The colour of each spot indicates the Z position of that spot. Spot projection to the XY plane at the bottom (magenta) highlights the XY positions of spots relative to the nucleus boundary (blue), that is, nucleus projection on the XY plane. The grid interval on the X and Y axis is ca. 18.3 µm. (D) Total count of mRNAs in each cell. The sample size (n) indicates the number of cells segmented in the microscope images. (E) Intracellular distribution of viral mRNAs. Spots in the 3D volume were projected on the XY plane, together with the cytoplasmic boundary (black solid line) and the nucleus (blue solid line). The grid interval indicates ca. 18.3 µm. (F) Spot segmentation within cells. The panel on the left shows the entire field-of-view (measures ∼294 µm), and the panel on the right shows the close-up of the area indicated in the red square (∼73 µm). Spots within the 2D nuclear projection are indicated in blue (nuclear); spots within the expanded nuclear boundary twofold from the centre of mass of the nucleus in 2D are indicated in orange (perinuclear); and spots outside this expansion are indicated in green (peripheral). (G) Distribution of spots on the Z axis in the three segments (nuclear, perinuclear and peripheral) presented in panel F. The red dashed line shows the median Z coordinates of the peripheral spots, used to indicate the cell baseline. The number in the histogram for nuclear spots (0.731) indicates the fraction of spots that were above the cell base line in the field-of-view presented in panel F. (H) Fraction of spots above the cell baseline in the nuclear segment. The black dots indicate the fraction obtained in each field-of- view, and the blue bar indicates the mean. (I) Fraction of spots within the 2D boundary of the nucleus. The sample size is the same as in panel D.

We localised the mRNAs in 3D and counted the total number of mRNAs in each cell (Fig. 1C). The median number of mRNAs per cell was 145, 231, 310 and 303 molecules for PB1, HA, NP and NS, respectively (Fig. 1D), in line with the theoretical prediction that the rate of mRNA production is partly determined by the gene length.

What came intriguingly was the intracellular distribution of mRNAs of the four segments. The intracellular distribution of mRNAs projected on the XY plane (Fig. 1C) indicated that the HA mRNAs were more abundant in the nucleus compared to the other three segments (Fig. 1E).

We wished to quantify the nuclear ratio of mRNAs in each cell. The mRNA count within the 2D nuclear projection is considered to be a good approximation for the number of mRNAs contained in the nucleus 20, 21. We validated this approximation for the lung epithelial A549 cells used in this study by looking at the intracellular Z positions of mRNAs: The mRNAs within the nuclear projection would have higher Z positions over the cell baseline, considering that the nucleus has a diameter of several microns, much thicker than the cytoplasm in the cell periphery. To this end, mRNAs were segmented into three regions in the 2D projection, that is, nuclear, perinuclear and peripheral, depending on the relative location to the nucleus (Fig. 1F). In contrast to the perinuclear (orange) and peripheral (green) mRNAs, mRNAs within the nuclear projection (blue) were distributed around the nuclear focal plane Z = 0 (Fig. 1G); and about 80% of mRNAs in the nuclear projection (blue) were above the cell baseline (Fig. 1H), defined by the median of the Z distribution of peripheral mRNAs (Fig. 1G, red dashed line). Thus, in this paper, we consider the spots within the nuclear 2D projection to be the nuclear fraction for the sake of simplicity.

With this approximation, about 30% of the PB1, NP and NS mRNAs were nuclear, while ∼70% of HA mRNAs were in the nuclear fraction (Fig. 1I), confirming the visual inspection that the HA mRNAs are more abundant in the nucleus at the early time point of infection (Fig. 1E). This result indicates that the velocity of viral mRNA export from the nucleus varies according to the viral segments. Note that our fluorescent probes could also detect complementary RNA (cRNA), an intermediate RNA that is produced for the viral genome replication, as it has the same polarity as viral mRNA. However, the fluorescent signals detected at 2 hours post-infection are deemed to be from the mRNAs, because cRNA production occurs much later than the mRNA synthesis 13, 22.

Mapping eight viral mRNAs in single cells by multiplex RNA FISH

Next, we wish to simultaneously quantify all the eight viral segments in situ. To this end, we applied MERFISH (multiplex error-robust RNA FISH) 23, an imaging-based transcriptome assay in which multiple rounds of probe hybridisation and imaging are performed to identify hundreds to thousands of mRNA species in situ by decoding binary fluorescent signals. For example, hundreds of distinct mRNAs can be encoded in 16-bit binary signals, in which case eight rounds of hybridisation and imaging would be required to read out the signals with two fluorophores 23. In our case of detecting eight influenza A virus segments, 6 bits were sufficient to accommodate all the eight mRNAs, with each code being separated by two or more binary flips from others (Fig. 2A). Thus, only three rounds of hybridisation and two-colour imaging were required. We sought for an economical way of performing sequential rounds of hybridisation and imaging, instead of setting up a custom- built, automated liquid pumping system used elsewhere 23. Inspired by an in vitro reconstitution assay of molecular motors 24, we assembled a flow cell on a slide glass, comprising two stripes of double-sided tape and the coverslip, with cells on the coverslip facing inwards (Fig. 2A). Liquids were introduced from the right open side and wicked from the left. During image acquisition, the two open sides were sealed with rubber cement. This easy implementation of flow cell allowed us to comfortably perform three rounds of hybridisation and two-colour imaging to read out the 6-bit binary tags; and by decoding the 6-bit encodings, the eight viral segments were identified in each cell (Fig. 2B). On average, about 100-200 molecules of each segment were contained per cell (Fig. 2C, top row). The difference in the total number of spots from the result presented in Fig. 1D is likely due to the batch effect (experimental variation).

Viral mRNA detection using multiplex RNA FISH.

(A) An overview of the assay. The viral mRNAs were hybridised with the encoding probes, and then the amplification probes were annealed (not shown in the diagram for simplicity). A flow cell was assembled in which the readout probes were hybridised and the cells were imaged in the three sequential rounds of hybridisation and imaging. Subsequently fluorescent signals were decoded according to the encodings listed on the right. (B) A representative mapping of identified viral mRNAs on the segmented cell and nuclear boundary in the entire field-of-view (left) and the close-up in the red square (right). The mesh grids indicate ∼34-µm interval on the left, and ∼17-µm interval on the right. (C) Total count of mRNAs in 204 single cells (top row) and the nuclear fraction (bottom row) for each segment. (D) Heatmap representation of mRNA counts in single cells. The viral segments are in rows, cells in columns. The cells were clustered column-wise to highlight the inequality in the abundance of transcripts carried per cell. The colour code for the absolute count is indicated in the vertical bar on the right. (E) A pair-wise analysis of the transcript count between segments.

The nuclear fraction of mRNAs varied between the viral segments (Fig. 2C, bottom row): While ∼30% of mRNAs were nuclear for PB1, NP and NS segments, about 70% of mRNAs remained in the nucleus for HA and NA segments; for the other segments (PB2, PA and M), around half of mRNAs were in the nucleus. This reiterates the idea that the velocity of viral mRNA export from the nucleus varies according to the viral segments. These results were confirmed by smFISH using probes for each segment that were synthesised economically in-house (Supplementary figure 1A, B and C).

Single-cell clustering according to the abundance of eight viral segments revealed that the cell population was heterogeneous in terms of the abundance of viral mRNAs (Fig. 2D). A pair-wise analysis of the abundance of viral segments indicated inequal distribution of viral mRNAs, regardless of the segments (Fig. 2E): Cells that carry abundant mRNAs from one segment tend to also carry abundant mRNAs from the other segments. These observations suggest that the viral transcription is susceptible to the stochastic variation in each infection 25, 26, which we subsequently used in our statistical model to estimate the nuclear export rate.

Estimating the nuclear export rate using population statistics

We suspected that the higher nuclear ratio of HA and NA mRNAs is due to the slower nuclear export rate for these mRNAs compared to other segments such as NP and NS. Thus, we wished to quantify the nuclear export rate of eight viral mRNA species from the observed nuclear-to-cytoplasmic distribution. To this end, we conceptualised a statistical model using the cell-to-cell variation in the abundance of viral transcripts of each segment (Fig. 3A). In this model, influenza viral particles are added to the cell population at time 0 h, and are allowed to undergo viral attachment to the cell surface, membrane fusion, and vRNP migration to the nucleus. The waiting time for these processes vary in each infection due to the stochastic nature of biochemical events 25, 27, and therefore, the onset of viral transcription varies from cell to cell in the population. In other words, each cell will have a varying duration for viral transcription by the time the observation is made at time 2 h (Fig. 3A, top). As a result, the population contains front runners in which the viral transcription was initiated early, and therefore abundant viral transcripts are carried; slow starters in which the transcription began late, containing fewer transcripts; and many other cells that lie along the spectrum between the two. The cell population at the single point in time, therefore, forms a trajectory of the time development of viral transcription and subsequent nuclear export on the scale of nuclear fraction against the abundance of transcripts (Fig. 3A, bottom): In a spontaneous transcription burst, the nascent transcripts emerge in the nucleus at the early stage of the burst, and they are exported to the cytoplasm as more transcripts are being synthesised 20.

Cell-to-cell heterogeneity for time-course inference.

(A) A theoretical diagram of how viral mRNAs emerge in single cells in the population. The font runners designate ones in which the viral transcription was initiated earlier than others, and the slow starters are ones that started transcription lately. The bottom plot shows the theoretical relationship between the nuclear ratio and the abundance of transcripts in which the front runners and slow starters (and many more along the spectrum in between) compose the clusters as indicated. (B) Scatter plots of nuclear fraction against the total count in each cell, derived from the dataset presented in Fig. 2C. (C) Time-course of the distribution of NP mRNA count per cell, measured by smFISH, at the indicated time points (top row); and the change in the nuclear fraction size, plotted against the total number of transcripts (bottom row). The sample size (n) indicates the number of cells observed at each time point.

Indeed, when the nuclear ratio of viral mRNA was plotted against the total quantity of mRNA molecules observed by multiplex RNA FISH in each cell, the plots formed a curve that originated from top left (i.e., nearly 100% of the nuclear ratio when a cell contained a few mRNA molecules), and descended as the number of mRNA molecules increased (30% to 70%, depending on the segments) (Fig. 3B and Supplementary figure 2).

We further validated this concept by performing smFISH measurements along the time-course with 40-min interval (Fig. 3C). A549 was infected with WSN at the M.O.I of ∼0.02. At each time point, the NP mRNAs were stained and quantified. At the beginning (0 and 40 min), the mRNA count was almost zero. As the mRNA count increased (80 min to 200 min post-infection) (Fig. 3C, top), the nuclear ratio against the total mRNA count derived from the single cells progressed along a curve, from the top left to the bottom right (Fig. 3C, bottom), supporting the concept of our statistical model.

The curvature of the trace appeared to vary between segments (Fig. 3B). We explored to see if the curvature could represent a proxy for the nuclear export rate. Thus, we formulated a simple kinetic model in which mRNAs are synthesised at the rate k in the nucleus, and are exported with the reaction constant ε (Fig. 4A). In this model, the molar concentrations of mRNAs in the nuclear and cytoplasmic compartments are described, respectively,

Kinetic model for estimating the nuclear export rate.

(A) A schematic diagram representing the production of viral transcripts in the nucleus and the export to the cytoplasm. The two spherical compartments are for the nucleus and the cytoplasm. Viral transcripts (magenta) are produced at a rate constant k in the nucleus, and exported at a rate constant ε to the cytoplasm. (B) The relationship between the nuclear fraction and the total count in the kinetic model, according to the formula presented in panel A. The parameter α determines the curvature of the plot. (C) The parameter fit to the observed datapoints. The parameter α was estimated by fitting the equation to the observed datapoints presented in Fig. 3B. The colour representation in each line is the same as the ones used in panel B, according to the parameter α. (D) Estimated nuclear export rate. The parameter ε was estimated from the three independent assays (indicated in black for each segment); the blue bar indicates the mean of the three measurements.

Omitting the degradation of mRNAs, the analytical solutions of these equations were:

Thus, the mRNA nuclear fraction φ was described as a function of total mRNA molecules Ntotal, where The nuclear fraction φ described in equation (5) is time-independent, and the parameter α determines the curvature of the line that the single cells would trace in the time-course (Fig. 4B). Thus, we estimated the parameter α by fitting to the observed data points derived at the time in which cells were fixed (Fig. 4C). By obtaining the synthesis rate k from the total number of mRNAs produced by the time of fixation (Fig. 2C, top row), the nuclear export rate ε for each segment was derived according to equation (6) (Fig. 4D). The nuclear export rate ε agreed with the historical classification of the early and late influenza A virus genes 4, 6, indicating that the timing of influenza A virus protein expression in the initial stage of infection is determined by the velocity of the nuclear export. Notably, the HA and NA mRNAs had the slowest nuclear export rate, suggesting that influenza A virus delays the viral antigen production by nuclear retention of viral transcripts.

Discussion

Based on the stochastic nature of influenza viral transcription studied by single-cell sequencing 25, 26, we introduced a quantitative framework for estimating the nuclear export rate of viral mRNAs to advanced our current understanding on the kinetics of viral gene expression. This was achieved by in situ localisation of eight influenza A virus segments simultaneously, coupled with the statistical model that allowed for estimating the nuclear export rate from the intracellular distribution of mRNAs at a single time point. This is based on the concept that the cells in the population are at various stages of transcription at the time the observation is made, due to the multiple stochastic processes the virus needs to overcome before it begins transcription. We demonstrated that the initiation of each viral transcription is concurrent in the eight viral segments; it is instead the nuclear export rate that controls the temporal protein expression at the initial phase of infection. The early production of the polymerase subunits (PB2, PB1 and PA) and nucleoprotein would facilitate the formation of vRNP, and hence the viral transcription and genome replication. On the other hand, pushing back the expression of viral antigens HA and NA would be beneficial for the virus to delay the host immune response against the infected cells in which the virus is being replicated.

Instead of time-series measurements, the estimation of mRNA export rate was made from a snapshot of in situ RNA localisation data taken at a single time point. We anticipate that the statistical framework for quantifying the nuclear export rate proposed in our study can be widely used and benefits the research area on the regulation of gene expression, especially since nuclear retention has been recognised as a point of control against the transcription burst 14. For example, it is applicable for the cell population in which de novo synthesis of mRNAs occurs concurrently, such as a stimuli-induced activation of immediate-early genes 21, 28. On the contrary, this approach is not appropriate for scenarios in which asynchronous transcription bursts are taking place randomly in each single cell 20, 29; cells at the late stage of transcription burst, having residual amount of transcripts in the cytoplasm 20, would obscure the time-course trajectory of the nuclear export.

The experiment in this study was conducted at the early stage of infection (up to 200 min post-infection). This conferred two advantages over investigating influenza viral mRNA export at the late stage of infection (e.g., 6-12 hours post- infection). First, our study allowed for the absolute quantification of viral transcripts at the single-molecule level in each cell, and thereby enabled us to formulate the model and fit the equation to the numerical data points. Otherwise at the late stage of infection, accurate quantification of mRNAs by the imaging- based single-molecule detection would not be feasible due to the high density of mRNAs. Second, at the late stage of infection, NS1 protein would be expressed in the cytoplasm. NS1 protein would then migrate into the nucleus and alter the nuclear export of viral mRNAs 30, obscuring the inherent nature of nuclear export of each viral segment. Thus, in our study, we elucidated the quantitative measure for the influenza viral mRNA nuclear export that is intrinsic to each viral segment.

Host factors that relay the influenza viral mRNAs to the nuclear pore remain elusive. NXF1 and CRM are two major cellular mRNA export pathways for the 5′-capped mRNAs 31. Of these two pathways, influenza A virus mRNAs, also carrying 5′ cap structure derived from the host nascent mRNA and the poly-A tail as a result of viral polymerase stuttering the uridine repeat, are dependent on the NXF1 pathway 16, 17. However, the eight viral segments show differential dependencies on this export pathway. The segments HA, M and NS is exported by NXF1 and UAP56, whereas PB2, PB1 and PA appear to be independent of the NXF1 pathway 17. The nuclear export rate derived in our study does not agree with the dependencies on NXF1. According to the previous study, both HA and NS are dependent on NXF1 and UAP56 16; however, in our study, HA and NS had significantly different nuclear export rate.

Two viral segments M and NS are subject to splicing, yielding unspliced and spliced transcripts. Our current system lacks the ability of discriminating between the unspliced and spliced forms of these segments; although we infer that the majority of NS segments that we detected in this study were in the unspliced form because we obtained similar result with probes exclusively detecting the unspliced form of NS (Supplementary figure 3). An observation that the unspliced form of M and NS dominates throughout the infection supports this 13. In either case, the RNA structure does not appear to accord with the NXF1 dependency, nor the nuclear export rate, either.

Our study suggests that the nuclear export of influenza viral transcripts is intrinsic to the nucleotide sequence of each segment. Recently, the human nuclear factor hnRNPAB was shown to associate with the nucleoprotein of influenza A virus, and inhibit the NXF1 pathway. It is tempting to speculate that the host factors (e.g., hnRNPAB) have differential binding affinities to each segment, thus determining the velocity of nuclear export (or the degree of nuclear export inhibition). It is currently unknown what factors binds to the early genes such as PB2, PB1 and PA. Further studies will be required to identify the host factors that bind to these transcripts for the nuclear export. Our study serves as a reference for the nuclear export rate of each segment, and thus sheds light to comprehensive understanding on the nuclear export of viral segments. Having revealed the nuclear retention of influenza viral transcripts at the unprecedented resolution, this study opens up a new research direction on how the nuclear retention is regulated and its implication in viral pathogenesis.

Materials and Methods

Cells and virus preparation

The human lung carcinoma cell line A549 was maintained in Dulbecco’s Modified Eagle’s Medium (DMEM) (Sigma, D5796) supplemented with 10% FBS (Nichirei Biosciences, 175012), 100 U/ml penicillin and 100 µg/ml streptomycin (Gibco, 15140-122) at 37°C in 5% CO2. Madin-Darby canine kidney (MDCK) cell line was maintained in Minimum Essential Medium (MEM) (Sigma, M4655) supplemented with 10% FBS, 100 U/ml penicillin and 100 µg/ml streptomycin at 37°C in 5% CO2. Influenza A virus strain WSN was propagated in MDCK cells in a 75 cm2 flask. The plaque-forming unit of the harvested virus was determined using MDCK cells grown confluent in a 6-well plate.

Coverslip preparation

An 18-mm round coverslip (0.17±0.005 mm thickness, Matsunami Glass) was used. For MERFISH, two straight sides were created by cutting two edges off the glass using a laser-wheel glass cutter (Nikken Dia, NC-X03 LASER 110°). The coverslip was rinsed in 70% ethanol and placed in a 12-well plate. A549 cells were seeded and attached onto the coverslip overnight. On the following day cells were rinsed twice with PBS and infected with WSN at the indicated multiplicity of infection in MEM for 2 hours (unless indicated otherwise in Fig. 3C) in the absence of FBS. After the infection, cells were fixed with 4% paraformaldehyde (Electron Microscopy Sciences, 15714-S) in PBS for 10 min and then rinsed three times with PBS. The coverslip was immersed in 70-75% ethanol and stored at −20°C until use.

smFISH

The probes used in Figs. 1, 3C, Supplementary figure 1 (probes for NP) and Supplementary figure 3 were obtained from LGC Biosearch Technologies. Stellaris RNA FISH Probe Designer (https://www. biosearchtech.com/stellaris-designer) was used to design the probes against the coding region of WSN mRNAs.

The coverslip was equilibrated in 1 ml of Wash buffer A (Stellaris) for >5 min in a 12-well plate prior to hybridisation. Cells were hybridised to 125 nM probes in Hybridization buffer (Stellaris) on a Hybrislip (Sigma, GBL712222) at 37°C overnight in a humidified chamber. On the following day the coverslip was rinsed twice in 1 ml of Wash buffer A (Stellaris), each for 30 min. Cytoplasmic stain (HCS Cellmask Green Stain, Invitrogen, H32714) was included in the second wash at the concentration of 50 ng/ml in Figs 1, 3C and Supplementary figure 3. Then the coverslip was rinsed in 1 ml of Wash buffer B (Stellaris) for 5 min. The coverslip was mounted in anti-fade medium containing DAPI (Vector Laboratories, H-1200) on a slide glass and sealed with nail polish.

For an economical reason and to allow for more flexible design of probes to increase the number of fluorophores attached per mRNA molecule, the probes used in Supplementary figure 1 (except the NP probes) were synthesised in-house (Supplementary figure 1A). (The NP probe used in Supplementary figure 1 was purchased from LGC Biosearch.) Candidate probe-binding sites in the coding region of each viral segment were identified using OligoMiner 32, and were further selected against the human RefSeq. The sequences were flanked by P5 (5′ AAT GAT ACG GCG ACC ACC GA 3′) and P7 (5′ CAA GCA GAA GAC GGC ATA CGA GAT 3′), with P7 at the 5′ end and P5 at the 3′ end of the probe sequence (Supplementary figure 1A). For the probes against HA, NA, M and NS, additional 20-nt sequences RS0332 (5′ GGG AGA ATG AGG TGT AAT GT 3′) and RS0406 (5′ GAT GAT GTA GTA GTA AGG GT 3′) 23 were inserted between P7 and the probe, the probe and P5, respectively (Supplementary figure 1A). These two insertions were used for additional probes to bind to increase the number of fluorophores per molecule. An oligonucleotide pool comprising the set of probe-binding sites with these additional nucleotide sequences was synthesised at Integrated DNA Technologies (IDT). The oligonucleotide pool was subjected to PCR amplification (98°C for 30 sec; 18 cycles of 98°C for 5 sec, 64°C for 10 sec and 72°C for 20 sec; and 72°C for 5 min) (Phusion High-Fidelity DNA Polymerase, NEB, M0530S) using the P7 and a T7-fused P5 primer (5′ TAA TAC GAC TCA CTA TAG GGA ATG ATA CGG CGA CCA CCG A 3′) (Supplementary figure 1A). RNA was synthesised using T7 RNA polymerase (HiScribe T7 High Yield RNA Synthesis Kit, NEB, E2040S). Then RNA was reverse- transcribed using a recombinant Moloney leukemia virus reverse transcriptase (GeneAce cDNA Synthesis Kit, Nippon Gene, 319-08881) with an ATTO550- conjugated P7 primer at the 5′ end. The single-stranded DNA probe was purified using a spin column (Zymo Research, D7010) after RNA digestion with a cocktail of RNase H and RNase A. The probe was further purified by Urea-PAGE to remove any residual RNA and the fluorescent-labelled primer.

The coverslip was equilibrated in 1 ml of 30% formamide buffer composed of 2×SSC and 30% deionised formamide (Sigma, F9037) in a 12-well plate for >5 min prior to hybridisation. Cells were hybridised with the probes at the concentration of 5.6 nM per probe in 100 µl Hybridization buffer composed of 2×SSC, 30% deionised formamide, 0.1% yeast tRNA (Invitrogen, 15401-011), 1% (v/v) RNase inhibitor (NEB, M0314S) and 10% (w/v) Dextran Sulfate (Merck, S4030) 23 on a Hybrislip at 37°C overnight in a humidified chamber. On the following day the coverslip was rinsed twice in 1 ml of 30% formamide buffer, each for 30 min. DAPI was included at 5 ng/ml in the second wash if additional staining was not performed with the external probes (for PB2, PB1 and PA). In this case the coverslip was rinsed in 1 ml of PBS after the DAPI staining, mounted in anti-fade medium (Vector Laboratories, H-1900) on a slide glass and sealed with nail polish. For staining HA, NA, M and NS with the extra probes, the coverslip was first equilibrated with 10% formamide buffer composed of 2×SSC, 10% deionised formamide and 0.05% (v/v) RNase inhibitor, and then hybridised with 10 nM each of probes for RS0332 and RS0406 23, both labelled with Quasar 570 (LGC Biosearch) at the 3′ end, in Stellaris hybridisation buffer (containing 10% formamide) at 37°C in the humidified chamber for 1 hour. Cells were washed in 10% formamide buffer twice, with DAPI included in the second wash. The coverslip was rinsed in PBS, mounted in the anti-fade medium (Vector Laboratories, H-1900), and sealed with nail polish.

Sequential hybridisation and imaging in the flow cell

An oligonucleotide pool for synthesising the encoding probes used in Fig. 2, comprising the target-binding sequence, readout tags compatible with the two-step amplification probes 23, 33, and the P5 and P7 PCR tags, were obtained from IDT. The target-binding sites were identified in the coding region of each segment using OligoMiner 32 and selected against human RefSeq. The readout tags RS0332, RS0343, RS0406, RS0255, RS0015 and RS038 23, 33 were inserted adjacent to the upstream or downstream of the target-binding sites for the 6-bit encoding of each segment (Fig. 2A). Probes were synthesised by PCR, T7 transcription and reverse-transcription as described above.

A coverslip was equilibrated in 1 ml of 30% formamide buffer (2×SSC and 30% deionised formamide) 23. Cells were hybridised with 1.6 µM encoding probes in Hybridisation buffer composed of 2×SSC, 30% deionised formamide, 0.1% yeast tRNA, 1% RNase inhibitor, 10% dextran sulfate 23 on a Hybrislip at 37°C overnight in a humidified chamber. Cells were washed twice in 1 ml of 30% formamide buffer at 37°C, and then equilibrated in 1 ml of Readout-probe hybridisation and wash buffer composed of 2×SSC, 10% deionised formamide and 20 units/ml Murine RNase inhibitor 23. The first and second amplification probes for RS0332, RS0255, RS0343, RS0406, RS0015 and RS0384, described previously 33, were annealed sequentially at 10 nM of each probe in Readout-probe hybridisation and wash buffer for 2 hours at 37°C on a Hybrislip. The coverslip was washed for 30 min twice in Readout-probe hybridisation and wash buffer at each of the first and second probe hybridisation. At the final wash, 0.10 µm green-fluorescent carboxyl polystyrene beads (Bangs Laboratories, FCDG002) were included (∼0.1% v/v) as a fiducial marker for image registration. The cells and beads were fixed with 3.2% paraformaldehyde in PBS for 10 min. The coverslip was rinsed in PBS three times and stored in 2×SSC supplemented with 40 units/ml RNase inhibitor at 4°C overnight.

On the following day a flow cell was assembled on a slide glass, composed of two parallel stripes of double-sided tape (0.23 mm thickness, ∼5 mm apart) and the coverslip on top, with cells facing inwards (Fig. 2A). The cells at the top and bottom end of the coverslip were scraped using a cell scraper before the assembly and any trace amount of liquid remaining on the exposed surface was removed with filter paper to ensure that the coverslip is firmly attached to the tape. The chamber was immediately flushed with 2×SSC to avoid cells drying up.

Readout-probe hybridisation and wash buffer was flushed in the chamber to equilibrate the cells. Readout probes were introduced at 10 nM in Readout-probe hybridisation and wash buffer and incubated at 37°C in a humidified chamber for 30 min. Subsequently, Readout-probe hybridisation and wash buffer was flushed to wash away the probes. Imaging buffer, composed of 2×SSC, 50 mM Tris-HCl pH 8, 10% glucose (Fujifilm Wako, 049-31165), 2 mM Trolox (Sigma, 238813), 0.5 mg/ml glucose oxidase (Sigma, G2133), 40 µg/ml catalase (Sigma, C30) and 0.1% RNase inhibitor 23, was flushed to the flow cell prior to the image acquisition. The flow cell was sealed at the both sides with Fixogum rubber cement ejected through a 200-µl pipette tip during the image acquisition.

After the imaging, the disulfide bond linking the probe and the fluorophore was cleaved with Cleavage buffer comprising 2×SSC and 50 mM TCEP (Sigma, 646547) for 15 min. The chamber was flushed with Readout-probe hybridisation and wash buffer to wash away the free fluorophores prior to introducing the readout probes for the next round. Readout probes used in this study are listed in Table 1.

Imaging platform

Images were acquired using a Nikon Ti2-E with a Nikon Plan Apo Lambda 60x (NA 1.40) (Fig. 1, Supplementary figure 3) or 100x (NA 1.45) (Fig. 3C) objective lens equipped with a Prime 95B 25MM back-illuminated sCMOS camera (Photometrics). Otherwise, an Olympus IX83 equipped with an Olympus UPlan XApo 60x (NA 1.42) objective lens and a cooled monochrome CCD camera (DP80, Olympus) was used (Fig. 2 and Supplementary figure 1).

Spot quantification

Fluorescent spots detected by smFISH were quantified using either FISH-QUANT 34 (Figs. 1, 3C and Supplementary figure 3) or a more recent Python package Big-FISH 35 (Supplementary figure 1). Cytoplasmic and nuclear segmentation was performed using Cellpose 36, 37. Objects that were on the boundary of the image was eliminated. Cytoplasmic regions with no nucleus therein due to the erroneous segmentation was also eliminated. The identification numbers assigned to each object by Cellpose were rearranged from scratch so that the cytoplasm and nucleus of the same cell correspond for the downstream analyses. In order to make the Cellpose mask readable in FISH-QUANT, the Cellpose mask were converted into x and y coordinates by tracing each object in the mask using the Moore neighbourhood edge-finding algorithm.

For MERFISH signal decoding, signals from the green-fluorescent carboxyl beads were used to register the images. Images at the corresponding field-of-view from the three rounds of imaging were registered using phase_cross_correlation from the Python package skimage.registration. The registered images were filtered using Laplacian-of-Gaussian implemented in Big-FISH, and the image intensity was manually normalised. Fluorescent signals were then decoded using PixelSpotDecoder in DetectPixels implemented in starfish (https://github.com/spacetx/starfish").

Data analyses and presentation

Microscopy images were presented using the Python package microfilm (https://github.com/guiwitz/microfilm) (Fig. 1B). Cell clustering was performed with the method centroid implemented in scipy.cluster.hierarchy.linkage from SciPy (https://scipy.org/) (Fig. 2D). The parameter α in equation (5) was estimated using curve_fit implemented in scipy.optimize from SciPy (Fig. 4C).

eLife Assessment

This important study combines virology experiments and mathematical modeling to determine the nuclear export rate of each of the eight RNA segments of the influenza A virus, leading to the proposal that a specific retention of mRNA within the nucleus delays the expression of antigenic viral proteins. The proposed model for explaining the differential rate of export is compelling, going beyond the state of the art, but the experimental setup is incomplete and would benefit from additional approaches. The insight so far is interesting, but because in the end it is left as an observation, the overall advance remains limited.

Reviewer #1 (Public Review):

The authors have approached the study of the mechanism by which the two more antigenic proteins of the influenza A virus, hemagglutinin (HA) and neuraminidase (NA), are expressed later during the infection. For this aim, they set an experimental approach consisting of a 2-hour-long infection at a multiplicity of infection of 2 with the viral strain WSN. They used cells from the lung carcinoma cell line A549. They used the FISH technique to detect the mRNAs in situ and developed an imaging-based assay for mathematically modeling and estimating the nuclear export rate of each of the eight viral segments. They propose that the delay in the expression of HA and NA is based on the retention of their mRNA within the nucleus.

The main strength of this work is the fact that the authors have studied a long-unaddressed mechanism in influenza A virus infectious cycle, as is the late expression of HA and NA, by creating a work flow including mRNA detection (FISH) plus mathematical calculations to arrive at a model, which additionally could be useful for general biological processes where transcription occurs in a burst-like manner. The weakness of this work in its present state is that in order to "quantify" the export rate of the transcripts, several assumptions regarding the viral infection are made without empirical data. It would greatly improve if more precise experiments could be performed and/or include demonstration of the assumptions made (i.e., synchronized infections, empirically demonstrating that cRNA production does not occur within the first 2 hours of infection, and/or separate transcription and replication, inhibiting RNA degradation during viral infection).

Reviewer #2 (Public Review):

In this study the authors developed a framework to investigate the export rates of Influenza viral RNAs translocating from the nucleus to the cytoplasm. This model suggests that the influenza virus may control gene expression at the RNA export level, namely, the retention of certain transcripts in the nucleus for longer times, allows the generation of other viral encoded proteins that are exported regularly, and only later on do certain mRNAs get exported. These encode proteins that alert the cell to the presence of viral molecules, hence keeping their emergence to very end, might help the virus to avoid detection as late as possible in the infection cycle.

The study is of limited scope. The notion that some mRNAs are retained in the nucleus after transcription is concluded early on from the FISH data. The model does not contribute much to the understanding and is mostly confirming the FISH data. The export rate is an ambiguous number and this part is not elaborated upon. One is left with more questions since no mechanistic knowledge emerges, and no additional experimentation is attempted to try drive to a deeper understanding.

Author response:

The following is the authors’ response to the original reviews.

We thank the editors and reviewers for the comments and suggestions on our manuscript. The main point that we wished to convey in this paper was the concept and the kinetic model that enabled the estimation of nuclear export rate from an image of single mRNAs localised in single cells. By studying the influenza viral transcripts with this model, we report the variation in the mRNA nuclear export rate of the eight viral segments. Of note, the hemagglutinin and neuraminidase mRNAs were the slowest among the eight segments in exiting the nucleus. We agree that the potential mechanism and the biological impact of this observation require further validation, as the reviewers pointed out. We revised our manuscript to describe these points separately (Lines 21-25, Abstract; Lines 86-91, Introduction; Lines 316-320, Results; Lines 372-381, Discussion). We also highlight below, the revisions that we made to address the specific points raised by the reviewers.

Influenza viral transcription

The authors used specific settings for their virology experiments and several assumptions regarding their mathematical modelling, so it's extremely important that the reader has the viral life cycle clearly understood before immersing themselves in the results. Thus, a detailed explanation of the viral life cycle, including the kinetics of each step, would be extremely helpful if included in the introduction section. Reviewer #1

We have included the molecular composition of influenza vRNP and the mechanism of viral transcription in the revised manuscript (Lines 46-53).

Line 45: "Eight viral RNA segments are transcribed by the same set of molecular machinery" (Ref. 7). What's known about the arrival of the viral RNA segments in the nucleus? Is it synchronized? The authors will understand that my concern is related to the fact that a differential arrival would indeed impact the transcription and export processes. Reviewer #1

The arrival of eight vRNPs in the nucleus is not synchronised, with each of the eight vRNPs arriving independently (Chou et al. PLOS Pathogens 2013) (Lakadamyali et al, PNAS 2003). This does not compromise our model, as our model estimates the export rate of each mRNA species individually (also please see our response in Model assumption below). This is included in the second paragraph of the Discussion section (Lines 390-400).

Model assumption

Even though I do not have the expertise to assess the authors' mathematical model, I do not doubt its robustness. Even so, I find some virological concerns related to the set-up of their experiments. According to what I understand, the authors performed non-synchronized 2 h-long infections with the WSN strain of influenza A virus. They did this to avoid cRNA production (and cross-reaction of the probes), which they claim to occur "much later than mRNA synthesis". Then they omit the degradation of the mRNAs for their model without giving an explanation for having done so. So, taking all these into account, it seems to me that too many assumptions are made without a strong argument. I understand that they are made in order to simplify their model, but I strongly consider that the model would gain strength if some of these events were experimentally considered. Thus, would it be possible to perform synchronized infections? Would it be possible to empirically demonstrate that cRNA production does not occur within the first 2 hours of infection and/or separate transcription and replication? Would it be possible to incorporate a degradation inhibitor of the mRNAs into their infections? If all these could be achieved, then the results coming out of the mathematical model would be enormously reinforced. Reviewer #1

* The study lacks experimental data that would help support the conclusions. For instance, perturbations are many times used to prove a point related to gene expression. An example for Fig. 2 for such an experiment could be to treat the cells with transcription inhibitors (e.g. DRB, 5,6-dichloro1-beta-D-ribofuranosylbenzimidazole). Preventing transcription leaves only mature RNAs in the nucleus, and then using this system one can compare the export rate of different RNAs. Reviewer #2

We agreed that the primary concern in our model was the assumption that the mRNA degradation could be omitted. Synchronised infection is not necessary; in fact, non-synchronised infection is preferred, as we explain later in our response. Additionally, the dominance of mRNA production over the cRNA production has been documented elsewhere. To address mRNA degradation and validate our model estimation, we performed a time-course measurement using baloxavir. Baloxavir efficiently blocks the viral transcription by inhibiting the nuclease activity in PA. DRB, suggested by the reviewer, allows influenza viral transcription and causes viral transcripts to accumulate in the nucleus for unknown mechanisms (Amorim et al. Traffic 2007 and our observation using smFISH, not shown). The additional experiment, now presented in Fig. 5 in the revised manuscript, indicated that the mRNA degradation is minimal, and the export rate estimated in our model and the time-course experiment agreed well for the HA segment. The experiment raised the possibility that the time-course measurement underestimates the export rate of transcripts that exit the nucleus rapidly, such as NP. A real-time imaging of single transcripts would be necessary to directly measure the true nuclear export rate; however, this is beyond the scope of our paper. The new result is now presented in Fig. 5, Supplementary figures 3 and 4, and in the main text (Lines 322-360). An alteration was also made in Line 286 to guide to Fig. 5. The Materials and Methods section was updated (Lines 478-482).

We note that our model does not require synchronised infection. Even under synchronised infection, such as incubating cells with the virus at 4°C to facilitate attachment and subsequently shifting to 37°C to allow viral entry, the inherent heterogeneity in vRNP migration to the nucleus still remains. This randomness does not compromise our model; rather, our model exploits this random arrival of each vRNP in each cell in the system. This variation, in turn, generates cells carrying varying amounts of transcripts, enabling the estimation of nuclear export rate. Importantly, more variation ensures the broader distribution of transcript levels, enabling more precise parameter fitting in our model. It is also important to note that our model does not require the correlation between segments. Our model estimates the export rate of each mRNA species individually. These important points were explained in the Discussion section (Lines 390-400).

* There is no concrete value given for the export rates and what they might mean biologically (e.g. time present/stuck in the nucleus) - Fig. 4D. This leaves the reader in the dark. Reviewer #2

The export rate lambda (previously denoted as k) in our model (Fig. 4) and the decay constant k in the time-course measurement (Fig. 5) represent the proportion of mRNAs exported from the nucleus in an infinitesimal time, defining the nuclear export rate. This has been clarified in the revised manuscript (Lines 314-316), with some alterations to make the parameter use more comprehensive.

- The Greek letter k previously used in Fig. 4 and the associated equations was consistently replaced with lambda to avoid the confusion with the parameter k that is subsequently used for the exponent decay in Fig. 5 in the revised manuscript.

- The Greek letter epsilon (previously used to represent export) was replaced with mu, slightly more common for representing the rate of transport.

- The term “velocity” was consistently replaced with “rate” in the context of the nuclear export (Lines 163, 215, 320, 441).

- The phrase “molar concentrations of mRNAs” was corrected for “molecules of mRNAs” (Line 282).

Also, we have now described our model in two sections: “Conceiving the model” and “Implementing a kinetic model to estimate the nuclear export rate” in the Result. The first section outlines the conceptual framework of the model, and the second focuses on its implementation and the parameter extraction (Lines 227 and 277).

Applicability of the model

Lines 27-29. "Our framework presented in this study can be widely used for investigating the nuclear retention of nascent transcripts produced in a transcription burst." In my opinion, this is the strongest point of the manuscript: developing a mathematical model to analyze nuclear export retention as a mechanism of protein expression control, which could lay the foundation for further biological processes. The authors revisit this idea in the Discussion section. However, which would be those processes for which the model could be helpful? I consider that a more conspicuous discussion on this topic would broaden the readers scope, a crucial point under the eLife scope. Reviewer #1

* Could this framework be used to quantify the nuclear export rate of cellular RNAs? According to the explanation in the Discussion, it would seem that this approach is limited to quantifying the export rate of influenza RNAs. Reviewer #2

Our model is not limited to the influenza virus infection. Our model is applicable for systems where transcription is initiated concurrently, such as when stimuli trigger the activation of a certain set of genes for transcription. Therefore, this makes it particularly valuable for quantifying the nuclear retention of mRNAs in a transcription burst. This point is reiterated in Line 383-390.

Potential mechanisms for differential nuclear export rate of viral segments

* There is no mechanistic insight in the study. The idea driven by this study is that gene expression is regulated by the RNA export rate. But how is that explained? Is there any molecular pathway or explanation for this model? If the transcripts are ready for export, why do the mRNAs stay inside the nucleus? One option to consider are the export factors. Viral RNAs are exported by different pathways as mentioned (line 362), or by TREX2 (Bhat P et al Nat Comm 2023). The data shows that there is no difference observed in the export rate of different pathways. How about knocking down an important export factor to show how this affects the export rates. Or the opposite, overexpress a certain factor, would this change the nucleus/cytoplasm distribution of the retained RNAs. Reviewer #2

As we discussed in the paper, we are beginning to consider that each viral segment has an intrinsic sequence that determines its nuclear export rate, because previous studies on the export factors does not fully explain the variation in the nuclear export rate observed in our study. As the reviewer suggested, a recent study (Bhat et al. Nature Communications 2023) exactly pointed out the internal sequence in the HA segment, aligning with our working hypothesis. This point is discussed and their work (Bhat et al. 2023) has been cited in the Discussion section in the revised manuscript (Lines 446-449).

Biological impact of the nuclear retention

The authors mention several times throughout the manuscript that the virus might use the nuclear retention of mRNA for HA and NA to postpone the expression of these antigenic molecules. At this point, I need to admit that a great question mark appeared in my mind, maybe related to the fact that some knowledge is lacking in my analysis. Lines 328-330: "On the other hand, pushing back the expression of viral antigens HA and NA would be beneficial for the virus to delay the host immune response against the infected cells in which the virus is being replicated." As I tend to understand, the host immune response recognizes HA and NA within the viral particle, if so and independently of the time that HA and Na arrive at the virus assembly step, the progeny' viral particles that are complete and extruded from the cells would be those awakening the host immunity response. If this is right, how would the delayed export of those proteins from the nucleus (and their late expression) be beneficial for delaying the immune response? I would appreciate an explanation for this point, and if I am wrong, then there could exist a relationship between nuclear export rate and the pathogenicity of different strains of influenza A virus. If so, could the authors challenge their model with additional viral strains showing a differential immune response pattern? A deeper analysis in this direction would greatly strengthen the message in their manuscript. Reviewer #1

* Is the timing of viral protein appearance in accordance with the time the mRNA is exported to the cytoplasm. It is logical that the first mRNA to go to the cytoplasm would be the first to become a protein. Can the authors show that nuclear retention of mRNA would push back the expression of the viral antigens HA and NA. Reviewer #2

Three types of immune reactions are being studied extensively. The first is the humoral immune response, where antibodies target the viral antigens HA and NA on the viral envelope, coating and inactivating the viral particles. The second is the cytotoxic T cell response. There is growing evidence that cytotoxic T cells react against NP, eliciting cross-reaction to broader range of influenza viral strains. This reaction is not specific to HA and NA, and antigens are processed in the cytoplasm and presented on the MHC. The third is antibody-dependent cellular cytotoxicity (ADCC), where antibodies recognise the viral proteins on the cellular surface (HA and NA) of infected cells, facilitating their elimination by the NK cells. Although protein translation may begin as soon as the first mRNA exits the nucleus, the virus may delay the peak of the antigen production and therefore, postpone the NK-mediated ADCC. This specific point, along with references to ADCC in influenza virus infection, has been clarified in the Discussion section (Lines 377-381).

Data analysis and presentation

Lines 99-101. "Viral mRNAs were detected as single diffraction-limited spots in the three-dimensional image stacks, allowing for absolute mRNA quantification (Fig. 1B)". What do the authors mean to say by "absolute mRNA quantification"? Do they refer to the total spots or the total mRNAs? Is it assumed that one spot corresponds to a single mRNA transcript? This is not clear at all for this reviewer, which could be the situation for a potential reader. Since it's the beginning of the story, this should be clearly stated in the manuscript. Reviewer #1

Each spot of fluorescent signal corresponds to a single molecule of viral mRNA. We quantified the absolute number of transcripts in each cell. This is clarified in the revised manuscript (Lines 104-106).

* Line 151: does the baseline change according to the RNA in question? The authors say that the "baseline is defined by the median of the Z distribution of peripheral mRNAs" - it seems that the number 0.731 refers only to one type of RNA (which is not mentioned at all not in the text and not in the legend). Reviewer #2

The baseline was set using the NP mRNAs in the cytoplasm because the NP mRNA showed the widest distribution across the cytoplasm (Line 157).

* Also, what is all the signal that is seen outside the marked cells in Fig. 2B? There seems to be significant background in the field, does this mean much false-positive in the multiplex FISH? If so, then how do the authors know that the staining inside the cells isn't to some degree non-specific? It would be necessary to back this up with some other type of quantitative assay like qRT-PCR. Reviewer #2

The cells were removed from the analysis if the cytoplasmic boundary touched any edge of the field-of-view, while the signals were recovered across the entire field-of-view. This is clarified in the figure legend (Lines 194-195).

Others

* The meaning and explanation for Figure 1H -are unclear. Rephrase and make the legend more reader friendly. Reviewer #2

We made alterations to the legend (Lines 132-134) and the relevant lines in the main text (Lines 148-151).

* Fig. 2E: Is this the total transcript count or only in the nucleus? Would it be possible to find some correlation between the segments if a pair-wise analysis is performed according to nuclear-cytoplasm distribution? Reviewer #2

The total counts are presented. This is clarified in the legend (Lines 199-200).

* Abstract -"A mathematical modelling indicated that the relationship between the nuclear ratio and the total count of mRNAs in single cells is dictated by a proxy for the nuclear export rate." - this sentence is very unclear. Reviewer #2

The sentence was removed in the revised manuscript (Line 21). This removal did not affect the overall meaning in the abstract. We also made an alteration to Line 279 that contained a similar phrase.

* The use of the word "acutely" (lines 16 and 35) is strange. Reviewer #2

They have been removed (now Lines 15, 33).

* Line 157 - "This result indicates that the velocity of viral mRNA export from the nucleus varies according to the viral segments." - not velocity, maybe timing. Reviewer #2

We consistently replaced “velocity” with “rate” (Lines 163, 215, 320, 441).

* Reference for line 41. Reviewer #2

A reference (Waker et al. Trends Microbiol. 2019) has been cited (Line 39).

* Reference for lines 105-106. Reviewer #2

The gene length of each segment was indicated in the sentence (Line 137).

* Line 264- why here is 0.02 M.O.I used compared to line 97 where 2 is used? Reviewer #2

We used M.O.I. of 0.02 to allow for spot quantification over longer periods of observation (Lines 269-270).

* NS1 is expressed at late infection times and might alter the nuclear export of viral mRNAs (line 352). Need to show that indeed it is not expressed in the experiments done here. Reviewer #2

It is not possible to definitely prove that NS1 is not expressed due to the sensitivity limitations. However, we minimised the its impact by investigating at the early time point (Lines 415416).

* Line 459- 30% formamide? Is this correct or should it be 10%? Reviewer #2

This is correct. The probes used were longer than the others for smFISH. Therefore, we washed away the probes with the stringent condition.

Version 2

Abstract

Abstract

Influenza A virus transcribes viral mRNAs from the eight segmented viral genome when it infects. The kinetics of viral transcription, nuclear export of viral transcripts, and their potential variation between the eight segments are poorly characterised. Here we introduce a statistical framework for estimating the nuclear export rate of each segment from a snapshot of in situ mRNA localisation. This exploits the cell-to-cell variation at a single time point observed by an imaging-based in situ transcriptome assay. Using our model, we revealed the variation in the mRNA nuclear export rate of the eight viral segments. Notably, the two influenza viral antigens hemagglutinin and neuraminidase were the slowest segments in the nuclear export, suggesting the possibility that influenza A virus uses the nuclear retention of viral transcripts to delay the expression of antigenic molecules. Our framework presented in this study can be widely used for investigating the nuclear retention of nascent transcripts produced in a transcription burst.

Introduction

Temporal gene expression is crucial for viral pathogens to proliferate in the host. Human exogenous retroviruses reverse-transcribe their RNA genome, insert it into the host chromatin and use host factors to minimise the viral protein production to evade the immune response and persist in the host (1). In contrast, influenza A virus infects the respiratory tract and produces thousands of viral progenitors in a single round of infection to spread in the surrounding epithelial cells by transcribing and replicating its RNA genome (2). Influenza A virus genome is segmented into eight single-stranded RNAs, and expression of each gene is essential for productive infection (3). Unlike the human exogenous retroviruses, influenza A virus does not seem to have a sophisticated mechanism of controlling the viral gene expression at the transcription level (4), such as transcription-factor binding or epigenetic modifications. However, it has been known for decades that hemagglutinin (HA) and neuraminidase (NA), the two influenza viral antigens targeted by the host immune response, are produced late during the course of infection (5–7), a mechanism of which is currently unclear.

The eight viral RNA segments are transcribed by the same set of molecular machinery (8), and thus the mechanism of influenza viral transcription does not seem to explain the temporal protein expression. The RNA-dependent RNA polymerase of influenza A virus, comprising PB1 (polymerase basic 1), PB2 (polymerase basic 2), and PA (polymerase acidic) subunits, is responsible for viral transcription. These subunits are associated with the partially hybridised 5′ and 3′ joint end of a viral genomic RNA segment (9). The oligomeric viral nucleoprotein (NP) binds to the rest of the intramolecular loop of the viral genomic RNA to form a viral ribonucleoprotein complex (vRNP) (10). The vRNPs are released into the host cytoplasm upon infection and migrate into the host nucleus for viral transcription (11, 12). The vRNP initiates the viral transcription in the nucleus by deriving a 5′-capped short RNA fragment from the host nascent RNA on the stalled RNA polymerase II (Pol II) (4). At the end of transcription, the viral polymerase adds a poly-A tail to the viral mRNA by stuttering on a short uridine repeat in the viral RNA template (13). Thus, eight viral segments sharing the common mechanism of viral transcription with no additional layers of transcriptional controls (e.g., binding of transcription factors), the transcription of eight influenza viral genes is likely to occur at the same rate across the segments that is determined by the collision of the vRNP and the stalled host Pol II.

Another possible factor that would determine the rate of transcription is the length of viral RNA template (14, 15), given that the duration of transcription is rate-limiting. However, this does not agree with the order of protein expression in the influenza virus infection: PB2 (2.4 kb), PB1 (2.3 kb) and PA (2.2 kb), the three longest viral genes, and NP (1.8 kb) are expressed first, followed by the expression of three segments HA (1.8 kb), NA (1.5 kb) and M (1.0 kb) (5, 7).

A third possibility is the nuclear retention of viral mRNAs. Nuclear retention of mRNAs has gained an attention as a point of control on the protein expression in the context of transcription burst (16). Previous studies reported the accumulation of M mRNAs in the nuclear speckles (17), NXF1-dependent nuclear export of HA, NA and M mRNAs (18, 19), and the nuclear export inhibition by the host factor hnRNPAB for M, HA, NP and NS mRNAs (20). It is becoming clear that the influenza viral segments show differential dependencies on the nuclear export factors; and this dependency is not accounted for by the RNA structure (i.e., whether they are intron-less, unspliced or spliced). Because an incomplete set of viral segments was studied in each of these studies, the influenza viral mRNA transport remains elusive.

In this paper, we introduce a statistical framework for systematically quantifying the nuclear transport across all the eight viral segments. We localised eight viral mRNAs simultaneously in single cells in situ. Intracellular distribution of viral mRNAs indicated that the nuclear export rate of viral transcripts varied according to the segments. A statistical model, exploiting cell-to-cell variation in the abundance of transcripts due to the stochastic nature of multiple virological processes prior to the onset of viral transcription, allowed for the estimation of nuclear export rate from the intracellular distribution of mRNAs observed at a single time point. With this model, we revealed the variation of the nuclear export rate in the eight viral segments. Notably, we show that hemagglutinin and neuraminidase mRNAs were the slowest in exiting the nucleus, suggesting the possibility that influenza A virus uses nuclear retention of mRNAs to delay the production of antigenic molecules in the course of viral infection. This study demonstrates that the nuclear export rate of nascent transcripts can be estimated from a snapshot of mRNAs localised in single cells. Our finding using this model serves as a reference for the nuclear export rate of influenza viral transcripts across the segments, shedding light on the comprehensive understanding on the nuclear export of viral transcripts.

Results

Diffraction-limited, single-colour imaging of viral mRNAs

First, we used single-molecule RNA FISH (smFISH) to study the viral transcription of PB1, HA, NP and NS. The lung carcinoma cell line A549 was infected with the WSN strain of influenza A virus at the multiplicity of infection (M.O.I.) of ∼2 for two hours, and the viral mRNAs were visualised with a tile of short oligonucleotide probes, each labelled with a single fluorophore (21) (Fig. 1A). Viral mRNAs were detected as single diffraction-limited spots in the three-dimensional image stacks. Each spot represents a single mRNA molecule (22), allowing for quantification of the absolute number of mRNA molecules in individual cells (Fig. 1B).

We localised the mRNAs in 3D and counted the total number of mRNAs in each cell (Fig. 1C). The median number of mRNAs per cell was 145, 231, 310 and 303 molecules for PB1 (2.3 kb), HA (1.8 kb), NP (1.8 kb) and NS (1.0 kb), respectively (Fig. 1D), in line with the theoretical prediction that the rate of mRNA production is partly determined by the gene length.

Viral mRNA detection by single-colour RNA FISH.

(A) A schematic diagram of mRNA labelling for the diffraction-limited imaging. Each mRNA was hybridised with a tile of oligonucleotide, each labelled with a single fluorophore at the 3′ end. (B) A representative field-of-view of the microscope image in which NP mRNAs were stained with Quasar 670. White, smFISH signal; blue, nucleus stained with DAPI. The red square indicates the area for the sequential close-up of the image. The signal intensity is shown on the same grey scale across the three panels, indicated on the right. (C) Spot detection in 3D. Spots are projected to either the XZ or YZ plane to indicate the distribution of spots in a typical 3D cell volume. The colour of each spot indicates the Z position of that spot. Spot projection to the XY plane at the bottom (magenta) highlights the XY positions of spots relative to the nucleus boundary (blue), that is, nucleus projection on the XY plane. The grid interval on the X and Y axis is ca. 18.3 µm. (D) Total count of mRNAs in each cell. The sample size (n) indicates the number of cells segmented in the microscope images. (E) Intracellular distribution of viral mRNAs. Spots in the 3D volume were projected on the XY plane, together with the cytoplasmic boundary (black solid line) and the nucleus (blue solid line). The grid interval indicates ca. 18.3 µm. (F) Spot segmentation within cells. The panel on the left shows the entire field-of-view (measures ∼294 µm), and the panel on the right shows the close-up of the area indicated in the red square (∼73 µm). Spots within the 2D nuclear projection are indicated in blue (nuclear); spots within the expanded nuclear boundary twofold from the centre of mass of the nucleus in 2D are indicated in orange (perinuclear); and spots outside this expansion are indicated in green (peripheral). (G) Distribution of spots on the Z axis in the three segments (nuclear, perinuclear and peripheral) presented in panel F. The red dashed line shows the median Z coordinates of the peripheral spots, used to indicate the baseline of spots. The number in the histogram for nuclear spots (0.731) indicates the fraction of spots that were above the baseline in the field-of-view presented in panel F. (H) Fraction of spots in the 2D nuclear segment that were above the baseline (the red dashed line in panel G, defined in the main text). Each dot indicate the fraction calculated in each field-of-view, and the blue bar indicates the mean. (I) Fraction of spots within the 2D boundary of the nucleus. The sample size is the same as in panel D.

What came intriguingly was the intracellular distribution of mRNAs of the four segments. The intracellular distribution of mRNAs projected on the XY plane (Fig. 1C) indicated that the HA mRNAs were more abundant in the nucleus compared to the other three segments (Fig. 1E).

We wished to quantify the nuclear ratio of mRNAs in each cell. The mRNA count within the 2D nuclear projection is considered to be a good approximation for the number of mRNAs contained in the nucleus (23, 24). We validated this approximation for the lung epithelial A549 cells used in this study by looking at the intracellular Z positions of mRNAs: The mRNAs within the nuclear projection are expected to have higher Z positions over those of other mRNAs in the cell periphery, considering that the nucleus has the height of several microns, much thicker than the cytoplasm in the cell periphery. To this end, mRNAs were segmented into three regions in the 2D projection, that is, nuclear, perinuclear and peripheral, depending on the distance from the nucleus (Fig. 1F). In contrast to the perinuclear (orange) and peripheral (green) mRNAs, mRNAs within the nuclear projection (blue) were distributed around the nuclear focal plane Z = 0 (Fig. 1G); and about 80% of mRNAs in the nuclear projection (blue) were above the baseline (Fig. 1H), defined by the median of the Z distribution of peripheral NP mRNAs (Fig. 1G, red dashed line). Thus, in this paper, we consider the spots within the nuclear 2D projection to be the nuclear fraction for the sake of simplicity.

With this approximation, about 30% of the PB1, NP and NS mRNAs were nuclear, while ∼70% of HA mRNAs were in the nuclear fraction (Fig. 1I), confirming the visual inspection that the HA mRNAs are more abundant in the nucleus at the early time point of infection (Fig. 1E). This result indicates that the rate of viral mRNA export from the nucleus varies according to the viral segments. Note that our fluorescent probes could also detect complementary RNA (cRNA), an intermediate RNA that is produced for the viral genome replication, as it has the same polarity as viral mRNA. However, the fluorescent signals detected at 2 hours post-infection are deemed to be from the mRNAs, because cRNA production occurs much later than the mRNA synthesis (15, 25).

Mapping eight viral mRNAs in single cells by multiplex RNA FISH

Next, we wish to simultaneously quantify all the eight viral segments in situ. To this end, we applied MERFISH (multiplex error-robust RNA FISH) (26), an imaging-based transcriptome assay in which multiple rounds of probe hybridisation and imaging are performed to identify hundreds to thousands of mRNA species in situ by decoding binary fluorescent signals. For example, hundreds of distinct mRNAs can be encoded in 16-bit binary signals, in which case eight rounds of hybridisation and imaging would be required to read out the signals with two fluorophores (26). In our case of detecting eight influenza A virus segments, 6 bits were sufficient to accommodate all the eight mRNAs, with each code being separated by two or more binary flips from others (Fig. 2A). Thus, only three rounds of hybridisation and two-colour imaging were required. We sought for an economical way of performing sequential rounds of hybridisation and imaging, instead of setting up a custom-built, automated liquid pumping system used elsewhere (26). Inspired by an in vitro reconstitution assay of molecular motors (27), we assembled a flow cell on a slide glass, comprising two stripes of double-sided tape and the coverslip, with cells on the coverslip facing inwards (Fig. 2A). Liquids were introduced from the right open side and wicked from the left. During image acquisition, the two open sides were sealed with rubber cement. This easy implementation of flow cell allowed us to comfortably perform three rounds of hybridisation and two-colour imaging to read out the 6-bit binary tags; and by decoding the 6-bit encodings, the eight viral segments were identified in each cell (Fig. 2B). On average, about 100-200 molecules of each segment were contained per cell (Fig. 2C, top row). The difference in the total number of spots from the result presented in Fig. 1D is likely due to the batch effect (experimental variation).

Viral mRNA detection using multiplex RNA FISH.

(A) An overview of the assay. The viral mRNAs were hybridised with the encoding probes, and then the amplification probes were annealed (not shown in the diagram for simplicity). A flow cell was assembled in which the readout probes were hybridised and the cells were imaged in the three sequential rounds of hybridisation and imaging. Subsequently fluorescent signals were decoded according to the encodings listed on the right. (B) A representative mapping of identified viral mRNAs on the segmented cell and nuclear boundary in the entire field-of-view (left) and the close-up in the red square (right). The mesh grids indicate ∼34-µm interval on the left, and ∼17-µm interval on the right. Cells were removed if their cytoplasmic boundaries touched any edge of the field-of-view. (C) Total count of mRNAs in 204 single cells (top row) and the nuclear fraction (bottom row) for each segment. (D) Heatmap representation of mRNA counts in single cells. The viral segments are in rows, cells in columns. The cells were clustered column-wise to highlight the inequality in the abundance of transcripts carried per cell. The colour code for the absolute count is indicated in the vertical bar on the right. (E) A pair-wise analysis of the total transcript counts per cell between segments.

The nuclear fraction of mRNAs varied between the viral segments (Fig. 2C, bottom row): While ∼30% of mRNAs were nuclear for PB1, NP and NS segments, about 70% of mRNAs remained in the nucleus for HA and NA segments; for the other segments (PB2, PA and M), around half of mRNAs were in the nucleus. This reiterates the idea that the rate of viral mRNA export from the nucleus varies according to the viral segments. These results were confirmed by smFISH using probes for each segment that were synthesised economically in-house (Supplementary figure 1A, B and C).

Single-cell clustering according to the abundance of eight viral segments revealed that the cell population was heterogeneous in terms of the abundance of viral mRNAs (Fig. 2D). A pair-wise analysis of the abundance of viral segments indicated inequal distribution of viral mRNAs, regardless of the segments (Fig. 2E): Cells that carry abundant mRNAs from one segment tend to also carry abundant mRNAs from the other segments. These observations suggest that the viral transcription is susceptible to the stochastic variation in each infection (28, 29), which we subsequently used in our statistical model to estimate the nuclear export rate.

Conceiving the model

We suspected that the higher nuclear ratio of HA and NA mRNAs is due to the slower nuclear export rate for these mRNAs compared to other segments such as NP and NS. Thus, we wished to quantify the nuclear export rate of eight viral mRNA species from the observed nuclear-to-cytoplasmic distribution. To this end, we conceived a statistical model using the cell-to-cell variation in the abundance of viral transcripts of each segment (Fig. 3A). In this model, influenza viral particles are added to the cell population at time 0 h, and are allowed to undergo viral attachment to the cell surface, membrane fusion, and vRNP migration to the nucleus. The waiting time for these processes vary in each infection due to the stochastic nature of biochemical events (12, 28, 30), and therefore, the onset of viral transcription varies from cell to cell in the population. In other words, each cell will have a varying duration for viral transcription by the time the observation is made at time 2 h (Fig. 3A, top). As a result, the population contains “front runners” in which the viral transcription was initiated early, and therefore abundant viral transcripts are carried; “slow starters” in which the transcription began late, containing fewer transcripts; and many other cells that lie along the spectrum between the two. The cell population at the single point in time, therefore, forms a trajectory of the time development of viral transcription and subsequent nuclear export on the scale of nuclear fraction against the abundance of transcripts (Fig. 3A, bottom): In a spontaneous transcription burst, the nascent transcripts emerge in the nucleus at the early stage of the burst, and they are exported to the cytoplasm as more transcripts are being synthesised (23).

Indeed, when the nuclear ratio of viral mRNA was plotted against the total quantity of mRNA molecules observed by multiplex RNA FISH in each cell, the plots formed a curve: It originated from top left (i.e., nearly 100% of the nuclear ratio when a cell contained a few mRNA molecules), and descended as the number of mRNA molecules increased (30% to 70%, depending on the segments) (Fig. 3B and Supplementary figure 2).

Cell-to-cell heterogeneity for time-course inference.

(A) A theoretical diagram of how viral mRNAs emerge in single cells in the population. The font runners designate ones in which the viral transcription was initiated earlier than others, and the slow starters are ones that started transcription lately. The bottom plot shows the theoretical relationship between the nuclear ratio and the abundance of transcripts in which the front runners and slow starters (and many more along the spectrum in between) compose the clusters as indicated. (B) Scatter plots of nuclear fraction against the total count in each cell, derived from the dataset presented in Fig. 2C. (C) Time-course of the distribution of total NP mRNA count per cell, measured by smFISH, at the indicated time points (top row); and the change in the nuclear fraction size, plotted against the total number of transcripts (bottom row). The sample size (n) indicates the number of cells observed at each time point.

We further validated this concept by performing smFISH measurements along the time-course with 40-min interval (Fig. 3C). A549 was infected with WSN at the M.O.I. of ∼0.02. This low M.O.I. infection allowed for spot quantification over longer timepoints without the spots becoming too crowded. At each time point, the NP mRNAs were stained and quantified. The total mRNA count was virtually zero at the beginning (0 and 40 min). As the mRNA count increased (80 min to 200 min post-infection) (Fig. 3C, top), the nuclear ratio against the total mRNA count obtained from the single cells progressed along a curve, from the top left to the bottom right (Fig. 3C, bottom), supporting the concept of our statistical model.

Implementing a kinetic model to estimate the nuclear export rate

The curvature of the trace varied between segments (Fig. 3B). We explored to see whether we could extract the nuclear export rate for each segment from the trace formed by the observed data. To this end, we formulated a simple kinetic model in which mRNAs are synthesised at the rate λ in the nucleus, and are exported with the reaction constant μ (Fig. 4A). In this model, the molecules of mRNAs in the nuclear and cytoplasmic compartments are described, respectively,

Omitting the degradation of mRNAs (see the next section), the analytical solutions of these equations were:

Thus, the mRNA nuclear fraction ρ was described as a function of total mRNA molecules Ntotal, where

Kinetic model for estimating the nuclear export rate.

(A) A schematic diagram representing the production of viral transcripts in the nucleus and the export to the cytoplasm. The two spherical compartments are for the nucleus and the cytoplasm. Viral transcripts (magenta) are produced at a rate constant λ in the nucleus, and exported at a rate constant μ to the cytoplasm. (B) The relationship between the nuclear fraction and the total count in the kinetic model, according to the formula presented in panel A. The parameter α determines the curvature of the plot. (C) The parameter fit to the observed datapoints. The parameter α was estimated by fitting the equation to the observed datapoints presented in Fig. 3B. The colour representation in each line is the same as the ones used in panel B, according to the parameter α. (D) Estimated nuclear export rate. The parameter μ was estimated from the three independent assays (indicated in black for each segment); the blue bar indicates the mean of the three measurements.

The nuclear fraction ρ described in equation (5) is time-independent, and the parameter α determines the curvature of the line that the single cells would trace in the time-course (Fig. 4B). Thus, we estimated the parameter α by fitting to the observed data points derived at the time in which cells were fixed (Fig. 4C). By obtaining the synthesis rate λ from the total number of mRNAs produced by the time of fixation (Fig. 2C, top row), the parameter μ for each segment was obtained according to equation (6) (Fig. 4D). The parameter μ represents the fraction of mRNAs in the nucleus that are exported within an infinitesimal time (i.e. the nuclear export rate). The result reveals the variation of nuclear export rate in the eight viral segments (Fig. 4D). The nuclear export rate μ agreed with the historical classification of the early and late influenza A virus genes (5, 7), suggesting the possibility that the timing of influenza A virus protein expression in the initial stage of infection is determined by the rate of the nuclear export.

Time-course study of viral mRNA degradation and nuclear export

We omitted the mRNA degradation in the kinetic model described in the previous section. To confirm that this simplification can be justified, we measured the viral mRNAs over 5 hours after halting the viral transcription (Fig. 5A). The influenza virus was allowed for transcription for 2 hours, after which the transcription was blocked by baloxavir (31), inhibiting the nuclease activity that resides in the viral polymerase subunit PA. We examined the HA and NP segments by smFISH, the two extreme examples that showed the highest and lowest nuclear export rate in our model, and found no reduction of viral mRNAs over the 5 hours after inhibiting the viral transcription (Fig. 5B and Supplementary figure 3).

Next, we estimated the mRNA export rate using the time-course measurement. The nuclear ratio of viral mRNAs was plotted at each time point, and the decay constant of the exponential decay (i.e. the rate of nuclear export, corresponding to the parameter μ in our model) was fitted against the nuclear ratio that declined over time (Fig. 5C and Supplementary figure 4). For the HA segment, the nuclear export rate estimated using all the time points (indicated with the blue line) was 0.243 h-1 (the mean of three replicates), in a good agreement with the one estimated in our model. On the other hand, for the NP segment, the nuclear export rate estimated from the time-course measurement was 0.631 h-1 (the mean of three replicates), which was much lower compared with our model estimation for the NP segment. We reason that the NP mRNAs are exported so rapidly that, in the time-course study, the nuclear ratio had already started to approach its lowest by the first time point following the baloxavir treatment, leading to the underestimation of the mRNA export rate. Using only the first two time points for the decay-constant fit (indicated with the orange line) did not make a significant difference (estimated rate 0.972 h-1 for the NP segment, the mean of three replicates). This highlights that our model enables the estimation of mRNA export rate from a single snapshot of mRNA distribution in cells, eliminating the need for time-course measurements that could potentially underestimate the export rate.

Time-course experiment for viral mRNAs.

(A) A schematic diagram of the time-course measurement. Cells were infected with the virus two hours prior to the baloxavir treatment for halting the viral transcription. Cells were fixed at one-hour interval thereafter, then subjected to single-molecule RNA FISH. (B) The total mRNA counts over time post baloxavir treatment for the HA (top) and NP (bottom) segments. (C) The change of mRNA fraction detected in the nucleus over time post baloxavir treatment. The exponential decay was fitted against the data using either all the datapoints (indicated with the blue line) or the first two timepoints (indicated with the orange line). The exponents (i.e. the nuclear export rates) estimated are shown on the panel with the corresponding colour.

Discussion

Utilising the stochastic nature of influenza viral transcription (28, 29), we introduced a quantitative framework for estimating the nuclear export rate of viral mRNAs to advanced our current understanding on the kinetics of viral gene expression. This was achieved by in situ localisation of eight influenza A virus segments simultaneously, coupled with the statistical model that allowed for estimating the nuclear export rate from the intracellular distribution of mRNAs at a single time point. This is based on the concept that the cells in the population are at various stages of transcription at the time the observation is made, due to the multiple stochastic processes the virus needs to overcome before it begins transcription (12). We propose that the initiation of each viral transcription is concurrent in the eight viral segments; it is instead the nuclear export rate that controls the temporal protein expression at the initial phase of infection. The early production of the polymerase subunits (PB2, PB1 and PA) and nucleoprotein would facilitate the formation of vRNP, and hence the viral transcription and genome replication. On the other hand, pushing back the peak of protein expression of viral antigens HA and NA on the cellular surface might be beneficial for the virus to delay the Fcγ-receptor-mediated, antibody-dependent cellular cytotoxicity (32, 33) against the infected cells in which the virus is being replicated.

Instead of time-series measurements, the estimation of mRNA export rate was made from a snapshot of in situ RNA localisation data taken at a single time point. We anticipate that the statistical framework for quantifying the nuclear export rate proposed in our study have broad applications beyond the influenza virus infection. This benefits the research area on the regulation of gene expression, especially since nuclear retention has been recognised as a point of control against the transcription burst (16). For example, it is applicable for the cell population in which de novo synthesis of mRNAs occurs concurrently, such as a stimuli-induced activation of immediate-early genes (24, 34). In our use case for the influenza virus infection, synchronised infection was not necessary, such as incubating cells with the virus at 4°C to facilitate attachment and subsequently shifting to 37°C to allow viral entry. The inherent heterogeneity in vRNP migration to the nucleus still remains (11, 12). This randomness does not compromise our model; rather, our model exploits this random arrival of each vRNP in each cell in the system. This variation, in turn, generates cells carrying varying amounts of transcripts, enabling the estimation of nuclear export rate. Importantly, more variation ensures the broader distribution of transcript levels, allowing for more precise parameter fitting in our model. It is also important to note that our model does not require the correlation between segments. Our model estimates the export rate of each mRNA species individually. On the contrary, this approach is not appropriate for scenarios in which transcription bursts occur repeatedly and randomly in each individual cell within the population (23, 35); cells at the late stage of transcription burst, having residual amount of transcripts in the cytoplasm (23), would obscure the time-course trajectory of the nuclear export.

The experiment in this study was conducted at the early stage of infection (up to 200 min post-infection). This conferred two advantages over investigating influenza viral mRNA export at the late stage of infection (e.g., 6-12 hours post-infection). First, our study allowed for the absolute quantification of viral transcripts at the single-molecule level in each cell, and thereby enabled us to formulate the model and fit the equation to the numerical data points. Otherwise at the late stage of infection, accurate quantification of mRNAs by the imaging-based single-molecule detection would not be feasible due to the high density of mRNAs. Second, at the late stage of infection, NS1 protein would be expressed in the cytoplasm. NS1 protein would then migrate into the nucleus and alter the nuclear export of viral mRNAs (36), obscuring the inherent nature of nuclear export of each viral segment. Thus, minimising the impact of NS1 protein in our study, we elucidated the quantitative measure for the influenza viral mRNA nuclear export that is intrinsic to each viral segment.

Host factors that relay the influenza viral mRNAs to the nuclear pore remain elusive. NXF1 and CRM are two major cellular mRNA export pathways for the 5′-capped mRNAs (37). Of these two pathways, influenza A virus mRNAs, also carrying 5′-cap structure derived from the host nascent mRNA and the poly-A tail as a result of viral polymerase stuttering the uridine repeat, are dependent on the NXF1 pathway (18, 19). However, the eight viral segments show differential dependencies on this export pathway. The segments HA, M and NS is exported by NXF1 and UAP56, whereas PB2, PB1 and PA appear to be independent of the NXF1 pathway (19). The nuclear export rate derived in our study does not agree with the dependencies on NXF1. According to the previous study, both HA and NS are dependent on NXF1 and UAP56 (18); however, in our study, HA and NS had significantly different nuclear export rate.

Two viral segments M and NS are subject to splicing, yielding unspliced and spliced transcripts. Our current system lacks the ability of discriminating between the unspliced and spliced forms of these segments; although we infer that the majority of NS segments that we detected in this study were in the unspliced form because we obtained similar result with probes exclusively detecting the unspliced form of NS (Supplementary figure 5). An observation that the unspliced form of M and NS dominates throughout the infection supports this (15). In either case, the RNA structure (whether spliced or unspliced) does not appear to accord with the NXF1 dependency, nor the nuclear export rate, either.

Recently, the human nuclear factor hnRNPAB was shown to associate with the nucleoprotein of influenza A virus, and inhibit the NXF1 pathway. It is tempting to speculate that the host factors (e.g., hnRNPAB) have differential binding affinities to each segment, thus determining the rate of nuclear export (or the degree of nuclear export inhibition). It is currently unknown what factors binds to the early genes such as PB2, PB1 and PA. Further studies will be required to identify the host factors that bind to these transcripts for the nuclear export.

We suggest that the nuclear export of influenza viral transcripts is intrinsic to the nucleotide sequence of each segment. A recent study revealed the internal sequence in the influenza virus HA mRNA that determined its susceptibility to the nuclear export via TREX-2 at the nuclear pore complex (38). This finding offers promising insights into the intrinsic sequences encoded in influenza viral transcripts. Our study serves as a reference for the nuclear export rate of each segment, and thus sheds light to comprehensive understanding on the nuclear export of viral segments. By revealing the nuclear retention of influenza viral transcripts at the unprecedented resolution, this study opens up a new research direction on how the nuclear retention is regulated across the segments and its implication in viral pathogenesis.

Materials and methods

Cells and virus preparation

The human lung carcinoma cell line A549 was maintained in Dulbecco’s Modified Eagle’s Medium (DMEM) (Sigma, D5796) supplemented with 10% FBS (Nichirei Biosciences, 175012), 100 U/ml penicillin and 100 µg/ml streptomycin (Gibco, 15140-122) at 37°C in 5% CO2. Madin-Darby canine kidney (MDCK) cell line was maintained in Minimum Essential Medium (MEM) (Sigma, M4655) supplemented with 10% FBS, 100 U/ml penicillin and 100 µg/ml streptomycin at 37°C in 5% CO2. Influenza A virus strain WSN was propagated in MDCK cells in a 75 cm2 flask. The plaque-forming unit of the harvested virus was determined using MDCK cells grown confluent in a 6-well plate.

Coverslip preparation

An 18-mm round coverslip (0.17±0.005 mm thickness, Matsunami Glass) was used. For MERFISH, two straight sides were created by cutting two edges off the glass using a laser-wheel glass cutter (Nikken Dia, NC-X03 LASER 110°). The coverslip was rinsed in 70% ethanol and placed in a 12-well plate. A549 cells were seeded and attached onto the coverslip overnight. On the following day cells were rinsed twice with PBS and infected with WSN at the indicated multiplicity of infection in MEM for 2 hours (unless indicated otherwise in Fig. 3C) in the absence of FBS. After the infection, cells were fixed with 4% paraformaldehyde (Electron Microscopy Sciences, 15714-S) in PBS for 10 min and then rinsed three times with PBS. The coverslip was immersed in 70-75% ethanol and stored at −20°C until use.

For the baloxavir treatment in Fig. 5, cells were infected with WSN for two hours as described above. Then the medium with the virus was removed, and 800 µl of DMEM (containing 10% FBS, 100 U/ml penicillin and 100 µg/ml streptomycin) with 100 nM baloxavir acid (Shionogi) was added to each well. Subsequently coverslips were prepared by fixing the cells at the intervals indicated in Fig. 5A.

smFISH

The probes used in Figs. 1, 3C, Supplementary figure 1 (probes for NP) and Supplementary figure 5 were obtained from LGC Biosearch Technologies. Stellaris RNA FISH Probe Designer (https://www.biosearchtech.com/stellaris-designer) was used to design the probes against the coding region of WSN mRNAs.

The coverslip was equilibrated in 1 ml of Wash buffer A (Stellaris) for >5 min in a 12-well plate prior to hybridisation. Cells were hybridised to 125 nM probes in Hybridization buffer (Stellaris) on a Hybrislip (Sigma, GBL712222) at 37°C overnight in a humidified chamber. On the following day the coverslip was rinsed twice in 1 ml of Wash buffer A (Stellaris), each for 30 min. Cytoplasmic stain (HCS Cellmask Green Stain, Invitrogen, H32714) was included in the second wash at the concentration of 50 ng/ml in Figs 1, 3C and Supplementary figure 5. Then the coverslip was rinsed in 1 ml of Wash buffer B (Stellaris) for 5 min. The coverslip was mounted in anti-fade medium containing DAPI (Vector Laboratories, H-1200) on a slide glass and sealed with nail polish.

For an economical reason and to allow for more flexible design of probes to increase the number of fluorophores attached per mRNA molecule, the probes used in Supplementary figure 1 (except the NP probes), 3 and 4 were synthesised in-house (Supplementary figure 1A). (The NP probe used in Supplementary figure 1 was purchased from LGC Biosearch.) Candidate probe-binding sites in the coding region of each viral segment were identified using OligoMiner (39), and were further selected against the human RefSeq. The sequences were flanked by P5 (5′ AAT GAT ACG GCG ACC ACC GA 3′) and P7 (5′ CAA GCA GAA GAC GGC ATA CGA GAT 3′), with P7 at the 5′ end and P5 at the 3′ end of the probe sequence (Supplementary figure 1A). For the probes against HA, NA, M and NS, additional 20-nt sequences RS0332 (5′ GGG AGA ATG AGG TGT AAT GT 3′) and RS0406 (5′ GAT GAT GTA GTA GTA AGG GT 3′) (26) were inserted between P7 and the probe, the probe and P5, respectively (Supplementary figure 1A). These two insertions were used for additional probes to bind to increase the number of fluorophores per molecule. An oligonucleotide pool comprising the set of probe-binding sites with these additional nucleotide sequences was synthesised at Integrated DNA Technologies (IDT). The oligonucleotide pool was subjected to PCR amplification (98°C for 30 sec; 18 cycles of 98°C for 5 sec, 64°C for 10 sec and 72°C for 20 sec; and 72°C for 5 min) (Phusion High-Fidelity DNA Polymerase, NEB, M0530S) using the P7 and a T7-fused P5 primer (5′ TAA TAC GAC TCA CTA TAG GGA ATG ATA CGG CGA CCA CCG A 3′) (Supplementary figure 1A). RNA was synthesised using T7 RNA polymerase (HiScribe T7 High Yield RNA Synthesis Kit, NEB, E2040S). Then RNA was reverse-transcribed using a recombinant Moloney leukemia virus reverse transcriptase (GeneAce cDNA Synthesis Kit, Nippon Gene, 319-08881) with an ATTO550-conjugated P7 primer at the 5′ end. The single-stranded DNA probe was purified using a spin column (Zymo Research, D7010) after RNA digestion with a cocktail of RNase H and RNase A. The probe was further purified by Urea-PAGE to remove any residual RNA and the fluorescent-labelled primer.

The coverslip was equilibrated in 1 ml of 30% formamide buffer composed of 2×SSC and 30% deionised formamide (Sigma, F9037) in a 12-well plate for >5 min prior to hybridisation. Cells were hybridised with the probes at the concentration of 5.6 nM per probe in 100 µl Hybridization buffer composed of 2×SSC, 30% deionised formamide, 0.1% yeast tRNA (Invitrogen, 15401-011), 1% (v/v) RNase inhibitor (NEB, M0314S) and 10% (w/v) Dextran Sulfate (Merck, S4030) (26) on a Hybrislip at 37°C overnight in a humidified chamber. On the following day the coverslip was rinsed twice in 1 ml of 30% formamide buffer, each for 30 min. DAPI was included at 5 ng/ml in the second wash if additional staining was not performed with the external probes (for PB2, PB1 and PA). In this case the coverslip was rinsed in 1 ml of PBS after the DAPI staining, mounted in anti-fade medium (Vector Laboratories, H-1900) on a slide glass and sealed with nail polish. For staining HA, NA, M and NS with the extra probes, the coverslip was first equilibrated with 10% formamide buffer composed of 2×SSC, 10% deionised formamide and 0.05% (v/v) RNase inhibitor, and then hybridised with 10 nM each of probes for RS0332 and RS0406 (26), both labelled with Quasar 570 (LGC Biosearch) at the 3′ end, in Stellaris hybridisation buffer (containing 10% formamide) at 37°C in the humidified chamber for 1 hour. Cells were washed in 10% formamide buffer twice, with DAPI included in the second wash. The coverslip was rinsed in PBS, mounted in the anti-fade medium (Vector Laboratories, H-1900), and sealed with nail polish.

Sequential hybridisation and imaging in the flow cell

An oligonucleotide pool for synthesising the encoding probes used in Fig. 2, comprising the target-binding sequence, readout tags compatible with the two-step amplification probes (26, 40), and the P5 and P7 PCR tags, were obtained from IDT. The target-binding sites were identified in the coding region of each segment using OligoMiner (39) and selected against human RefSeq. The readout tags RS0332, RS0343, RS0406, RS0255, RS0015 and RS038 (26, 40) were inserted adjacent to the upstream or downstream of the target-binding sites for the 6-bit encoding of each segment (Fig. 2A). Probes were synthesised by PCR, T7 transcription and reverse-transcription as described above.

A coverslip was equilibrated in 1 ml of 30% formamide buffer (2×SSC and 30% deionised formamide) (26). Cells were hybridised with 1.6 µM encoding probes in Hybridisation buffer composed of 2×SSC, 30% deionised formamide, 0.1% yeast tRNA, 1% RNase inhibitor, 10% dextran sulfate (26) on a Hybrislip at 37°C overnight in a humidified chamber. Cells were washed twice in 1 ml of 30% formamide buffer at 37°C, and then equilibrated in 1 ml of Readout-probe hybridisation and wash buffer composed of 2×SSC, 10% deionised formamide and 20 units/ml Murine RNase inhibitor (26). The first and second amplification probes for RS0332, RS0255, RS0343, RS0406, RS0015 and RS0384, described previously (40), were annealed sequentially at 10 nM of each probe in Readout-probe hybridisation and wash buffer for 2 hours at 37°C on a Hybrislip. The coverslip was washed for 30 min twice in Readout-probe hybridisation and wash buffer at each of the first and second probe hybridisation. At the final wash, 0.10 µm green-fluorescent carboxyl polystyrene beads (Bangs Laboratories, FCDG002) were included (∼0.1% v/v) as a fiducial marker for image registration. The cells and beads were fixed with 3.2% paraformaldehyde in PBS for 10 min. The coverslip was rinsed in PBS three times and stored in 2×SSC supplemented with 40 units/ml RNase inhibitor at 4°C overnight.

On the following day a flow cell was assembled on a slide glass, composed of two parallel stripes of double-sided tape (0.23 mm thickness, ∼5 mm apart) and the coverslip on top, with cells facing inwards (Fig. 2A). The cells at the top and bottom end of the coverslip were scraped using a cell scraper before the assembly and any trace amount of liquid remaining on the exposed surface was removed with filter paper to ensure that the coverslip is firmly attached to the tape. The chamber was immediately flushed with 2×SSC to avoid cells drying up.

Readout-probe hybridisation and wash buffer was flushed in the chamber to equilibrate the cells. Readout probes were introduced at 10 nM in Readout-probe hybridisation and wash buffer and incubated at 37°C in a humidified chamber for 30 min. Subsequently, Readout-probe hybridisation and wash buffer was flushed to wash away the probes. Imaging buffer, composed of 2×SSC, 50 mM Tris-HCl pH 8, 10% glucose (Fujifilm Wako, 049-31165), 2 mM Trolox (Sigma, 238813), 0.5 mg/ml glucose oxidase (Sigma, G2133), 40 µg/ml catalase (Sigma, C30) and 0.1% RNase inhibitor (26), was flushed to the flow cell prior to the image acquisition. The flow cell was sealed at the both sides with Fixogum rubber cement ejected through a 200-µl pipette tip during the image acquisition.

After the imaging, the disulfide bond linking the probe and the fluorophore was cleaved with Cleavage buffer comprising 2×SSC and 50 mM TCEP (Sigma, 646547) for 15 min. The chamber was flushed with Readout-probe hybridisation and wash buffer to wash away the free fluorophores prior to introducing the readout probes for the next round. Readout probes used in this study are listed in Table 1.

Imaging platform

Images were acquired using a Nikon Ti2-E with a Nikon Plan Apo Lambda 60x (NA 1.40) (Fig. 1, Supplementary figure 5) or 100x (NA 1.45) (Fig. 3C) objective lens equipped with a Prime 95B 25MM back-illuminated sCMOS camera (Photometrics). Otherwise, an Olympus IX83 equipped with an Olympus UPlan XApo 60x (NA 1.42) objective lens and a cooled monochrome CCD camera (DP80, Olympus) was used (Fig. 2 and Supplementary figure 1-4).

Spot quantification

Fluorescent spots detected by smFISH were quantified using either FISH-QUANT (41) (Figs. 1, 3C and Supplementary figure 5) or a more recent Python package Big-FISH (42) (Supplementary figure 1-4). Cytoplasmic and nuclear segmentation was performed using Cellpose (43, 44). Objects that were on the boundary of the image was eliminated. Cytoplasmic regions with no nucleus therein due to the erroneous segmentation was also eliminated. The identification numbers assigned to each object by Cellpose were rearranged from scratch so that the cytoplasm and nucleus of the same cell correspond for the downstream analyses. In order to make the Cellpose mask readable in FISH-QUANT, the Cellpose mask were converted into x and y coordinates by tracing each object in the mask using the Moore neighbourhood edge-finding algorithm.

For MERFISH signal decoding, signals from the green-fluorescent carboxyl beads were used to register the images. Images at the corresponding field-of-view from the three rounds of imaging were registered using phase_cross_correlation from the Python package skimage.registration. The registered images were filtered using Laplacian-of-Gaussian implemented in Big-FISH, and the image intensity was manually normalised. Fluorescent signals were then decoded using PixelSpotDecoder in DetectPixels implemented in starfish (https://github.com/spacetx/starfish).

Data analyses and presentation

Microscopy images were presented using the Python package microfilm (https://github.com/guiwitz/microfilm) (Fig. 1B). Cell clustering was performed with the method centroid implemented in scipy.cluster.hierarchy.linkage from SciPy (https://scipy.org/) (Fig. 2D). The parameter α in equation (5) (Fig. 4C) and the exponents in Fig. 5C and Supplementary figure 4 were estimated using curve_fit implemented in scipy.optimize from SciPy.